\subsection*{Step-by-Step Semantic Reasoning with Embeddings}

\paragraph{1. What do we mean by \textquotedblleft semantic\textquotedblright?}
Semantic refers to \emph{meaning}—not spelling, not structure, but what a word or concept represents in context.

\textbf{Example:}
\begin{itemize}
	\item The words \texttt{apple} and \texttt{banana} are semantically close because both refer to fruits.
	\item The words \texttt{apple} and \texttt{laptop} may be semantically related only because of the company Apple, not in the food sense.
\end{itemize}
So when we say two words are semantically similar, we mean: \emph{“they appear in similar contexts or serve similar roles in language.”}

\paragraph{2. What is an embedding?}
An \emph{embedding} is a vector representation of a concept (word, sentence, etc.) that captures its semantic context in numerical form.

\textbf{Example:} In a model like Word2Vec:
\begin{align*}
	\texttt{cat} &= [0.5, -1.2, 0.3, \dots] \\\\
	\texttt{dog} &= [0.52, -1.15, 0.29, \dots]
\end{align*}

\paragraph{3. What does the subtraction $A - B$ mean in this context?}
When we compute $A - B$ for word vectors, we are finding the difference in meaning between two concepts.

\textbf{Example:}
\begin{align*}
	A &= \texttt{king} = [1.0, 0.8, 0.5, \dots] \\\\
	B &= \texttt{man} = [0.5, 0.3, 0.2, \dots] \\\\
	A - B &= \texttt{king} - \texttt{man} = [0.5, 0.5, 0.3, \dots]
\end{align*}
This result represents what distinguishes a \texttt{king} from a \texttt{man}—ideas like royalty or status.

Now consider:
\[
\texttt{woman} = [0.48, 0.32, 0.22, \dots]
\]
Then:
\[
\texttt{queen} \approx \texttt{king} - \texttt{man} + \texttt{woman}
\]

\paragraph{4. Where does this semantic encoding come from?}
These patterns are learned from massive language corpora.

\textbf{Example:}
If the model frequently sees:
\begin{itemize}
	\item “The \texttt{king} rules the kingdom.”
	\item “The \texttt{queen} rules the kingdom.”
	\item “The \texttt{man} is tall.”
	\item “The \texttt{woman} is tall.”
\end{itemize}
It learns that \texttt{king} and \texttt{queen} are used in parallel roles.

\[
\texttt{king} - \texttt{man} \approx \texttt{queen} - \texttt{woman}
\]

\paragraph{5. Summary Table}

\begin{table}[h!]
	\centering
	\begin{tabular}{|l|l|p{6cm}|}
		\hline
		\textbf{Concept} & \textbf{Meaning} & \textbf{Example} \\\hline
		\textbf{Semantic} & The meaning of a word or concept & \texttt{cat} and \texttt{dog} are semantically similar \\\hline
		\textbf{Embedding} & A numeric vector representing meaning & \texttt{cat} = $[0.5, -1.2, 0.3, \dots]$ \\\hline
		\textbf{Subtraction $A - B$} & Captures how $A$ differs from $B$ & \texttt{king} $-$ \texttt{man} $\approx$ ``royalty'' \\\hline
		\textbf{Semantic shift} & The difference in context or role & \texttt{woman} $+$ (\texttt{king} $-$ \texttt{man}) $\approx$ \texttt{queen} \\\hline
		\textbf{Encoding source} & Learned from language use & Learned by observing how words are used in similar roles \\\hline
	\end{tabular}
	\caption{Summary of semantic vector reasoning}
\end{table}

